---
title: "Succession classes"
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
```


Here we explore succession class over/under representation for 3 fire-dependent Biophysical Settings (BpS).  Each section will include:

* Chart of succession classes
* Quick summary
* Link to full BpS description

## Northeastern Interior Dry-Mesic Oak Forest

```{r dmof setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}

bps_scls_3 <- read.csv("data/bpsScls3.csv") %>%
  group_by(Var1) %>%
  mutate(total.count = sum(Freq)) %>%
  ungroup() %>%
  dplyr::filter(dense_rank(desc(total.count)) < 7) %>%
  dplyr::select(c("BpS_Name", "refLabel",  "currentPercent", "refPercent")) %>%
  pivot_longer(
    cols = c(`refPercent`, `currentPercent`), 
    names_to = "refCur", 
    values_to = "Percent"
  )


# order classes
bps_scls_3$refLabel <- factor(bps_scls_3$refLabel, levels= c(
  'Developed',
  'Agriculture',
  'Uncharacteristic exotic',
  'Uncharacteristic native',
  'E-Mixed mesophytic forests',
  'D-Oak and pine forest',
  'C-Oak savanna and woodlands',
  'B-Oak regeneration',
  'A-Herbaceous'
))

# reverse
factor(bps_scls_3$refLabel, levels = rev(levels(bps_scls_3$refLabel)))
```


```{r dmof chart, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=9}


sclasplot <-
  ggplot(bps_scls_3, aes(fill=factor(refCur), y=Percent, x=refLabel)) + 
  geom_col(width = 0.8, position = position_dodge()) +
  coord_flip() +
  scale_x_discrete(limits = (levels(bps_scls_3$refLabel))) +
  labs(
    title = "Succession Classes past and present",
    caption = "\nData from landfire.gov.",
    x = "",
    y = "Percent")+
  theme_minimal(base_size = 14)+
  theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot") +
  scale_fill_manual(values = c("#3d4740", "#32a852" ), # present (grey), historical (green)
                    name = " ", 
                    labels = c("Present",
                               "Past")) +
  theme(panel.spacing = unit(.05, "lines"))


sclasplotDMO <- sclasplot %+% subset(bps_scls_3, BpS_Name %in% c("Northeastern Interior Dry-Mesic Oak Forest"))


sclasplotDMO

```

Summary of succession classes past and present: Northeastern Interior Dry-Mesic Oak Forest

* Historically the Oak-Savanna and Woodlands would have been dominant, occupying almost 50% of this BpS, followed by the Oak and Pine Forest succession class at almost 30%.  
* The dominant succession classes were maintained by frequent low-severity fires.
* Today this BpS is dominated by the Mixed Mesophytic class at ~45%

**Download full description [here](https://github.com/rswaty/Monongahela/raw/main/assets/13030_61_63_64_65_66.docx)**

*Please read full description.  Succession class labels above greatly simplified.*

## Central and Southern Appalachian Montane Oak Forest

```{r camof setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}

bps_scls_3 <- read.csv("data/bpsScls3.csv") %>%
  group_by(Var1) %>%
  mutate(total.count = sum(Freq)) %>%
  ungroup() %>%
  dplyr::filter(dense_rank(desc(total.count)) < 7) %>%
  dplyr::select(c("BpS_Name", "refLabel",  "currentPercent", "refPercent")) %>%
  pivot_longer(
    cols = c(`refPercent`, `currentPercent`), 
    names_to = "refCur", 
    values_to = "Percent"
  )


# order classes
bps_scls_3$refLabel <- factor(bps_scls_3$refLabel, levels= c(
  'Developed',
  'Agriculture',
  'Uncharacteristic exotic',
  'Uncharacteristic native',
  'E-Mature closed canopy deciduous',
  'D-Mature open canopy deciduous',
  'C-Med. height open canopy deciduous',
  'B-Med. height closed canopy deciduous',
  'A-Oak regeneration'
))

# reverse
factor(bps_scls_3$refLabel, levels = rev(levels(bps_scls_3$refLabel)))

```





```{r camof chart, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=9}



sclasplot <-
  ggplot(bps_scls_3, aes(fill=factor(refCur), y=Percent, x=refLabel)) + 
  geom_col(width = 0.8, position = position_dodge()) +
  coord_flip() +
  scale_x_discrete(limits = (levels(bps_scls_3$refLabel))) +
  labs(
    title = "Succession Classes past and present",
    caption = "\nData from landfire.gov.",
    x = "",
    y = "Percent")+
  theme_minimal(base_size = 14)+
  theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot") +
  scale_fill_manual(values = c("#3d4740", "#32a852" ), # present (grey), historical (green)
                    name = " ", 
                    labels = c("Present",
                               "Past")) +
  theme(panel.spacing = unit(.05, "lines"))


sclasplotCAMOF <- sclasplot %+% subset(bps_scls_3, BpS_Name %in% c("Central and Southern Appalachian Montane Oak Forest"))


sclasplotCAMOF
```

Summary of succession classes past and present: Central and Southern Appalachian Montane Oak Forest

* Historically the more fire-dependent succession classes were dominant, cumulatively representing ~60% of this BpS.   
* LANDFIRE data suggests that classes A, B and C are missing on the landscape today.
* Today this BpS is dominated by the Mature closed canopy deciduous class at > 80%

**Download full description [here](https://github.com/rswaty/Monongahela/raw/main/assets/13200_48_57_61.docx)**

*Please read full description.  Succession class labels above greatly simplified.*

## Central Appalachian Dry Oak-Pine Forest

```{r cadop setup, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}

bps_scls_3 <- read.csv("data/bpsScls3.csv") %>%
  group_by(Var1) %>%
  mutate(total.count = sum(Freq)) %>%
  ungroup() %>%
  dplyr::filter(dense_rank(desc(total.count)) < 7) %>%
  dplyr::select(c("BpS_Name", "refLabel",  "currentPercent", "refPercent")) %>%
  pivot_longer(
    cols = c(`refPercent`, `currentPercent`), 
    names_to = "refCur", 
    values_to = "Percent"
  )


# order classes
bps_scls_3$refLabel <- factor(bps_scls_3$refLabel, levels= c(
  'Developed',
  'Agriculture',
  'Uncharacteristic exotic',
  'Uncharacteristic native',
  'E-mature closed canopy deciduous/conifers',
  'D-Mature open canopy deciduous',
  'C-Med. height open canopy deciduous',
  'B-Med. height closed canopy deciduous',
  'A-Oak regeneration'
))

# reverse
factor(bps_scls_3$refLabel, levels = rev(levels(bps_scls_3$refLabel)))

```





```{r cadop chart, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=9}



sclasplot <-
  ggplot(bps_scls_3, aes(fill=factor(refCur), y=Percent, x=refLabel)) + 
  geom_col(width = 0.8, position = position_dodge()) +
  coord_flip() +
  scale_x_discrete(limits = (levels(bps_scls_3$refLabel))) +
  labs(
    title = "Succession Classes past and present",
    caption = "\nData from landfire.gov.",
    x = "",
    y = "Percent")+
  theme_minimal(base_size = 14)+
  theme(plot.caption = element_text(hjust = 0, face= "italic"), #Default is hjust=1
        plot.title.position = "plot", #NEW parameter. Apply for subtitle too.
        plot.caption.position =  "plot") +
  scale_fill_manual(values = c("#3d4740", "#32a852" ), # present (grey), historical (green)
                    name = " ", 
                    labels = c("Present",
                               "Past")) +
  theme(panel.spacing = unit(.05, "lines"))


sclasplotCADOP <- sclasplot %+% subset(bps_scls_3, BpS_Name %in% c("Central Appalachian Dry Oak-Pine Forest"))


sclasplotCADOP
```

Summary of succession classes past and present: Central Appalachian Dry Oak-Pine Forest

* The Oak-regeneration and Mature classes are mapped by LANDFIRE as missing on the landscape today. HIstorically those two classes covered roughly 12% and 42% historically
* American Chestnut was an important part of this BpS (and possibly others).
* There is also an underrepresentation of the E-mature closed canopy deciduous/conifers class which is noted to be diverse in the description.


**Download full description [here](https://github.com/rswaty/Monongahela/raw/main/assets/13690_60_61_63_64_65_66.docx)**

*Please read full description.  Succession class labels above greatly simplified.*





